### 2009-06-01 paper.tex
Edits....

We describe a ``transducer composition'' algorithm for extending
pairwise statistical models of RNA structural evolution to models of
multiple sequences related by a phylogenetic tree.  This algorithm
draws on formal models of computational linguistics as well as the
1985 protosequence algorithm of David Sankoff.  The output of the
composition algorithm is a multiple-sequence stochastic context-free
grammar.  We describe dynamic programming algorithms, which are robust
to null cycles and empty bifurcations, for parsing this grammar.
Example applications include structural alignment of non-coding RNAs,
propagation of structural information from an
experimentally-characterized sequence to its homologs, and inference
of the ancestral structure of a set of diverged RNAs.

@@

A number of leading methods for bioinformatics analysis of noncoding RNA
 use probabilistic grammars as models for pairwise alignments.
We show that any such pairwise grammar can be extended to an entire phylogeny,
 by considering the pairwise grammar as a machine (a ``transducer'') that locally transforms one RNA structure into another
 (thereby modeling a single parent-child relationship in the tree).
In addition to current applications such as genefinding, homology detection, alignment and secondary structure prediction,
 this should permit statistical reconstruction of RNA genes ancestral to many present-day sequences.
We describe algorithms, software implementations, and an elementary benchmark wherein
 the complexity and accuracy of three-way RNA reconstruction is measured as the length of one branch is varied.
In the Discussion we consider how the three-way RNA alignment-reconstruction-folding algorithm, which is currently very expensive computationally,
 might be made more efficient so that larger phylogenies could be considered.

@@

% (second part: multiple)
The second half of the phylogenetic modeling problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

@@

The main contributions of this paper are
(1) an algorithm that transforms a phylogenetic ensemble of pair
grammars, representing models on branches of a phylogenetic tree,
into a coherent, multiple-sequence SCFG,
(2) dynamic programming (DP) algorithms for performing inference under
this multiple-sequence SCFG, and
% (which appear to converge even if the SCFG contains null bifurcations and null cycles);
(3) freely-available software implementing algorithms (1) and (2) for
the simplified case of a three-branch star-topology tree.
While the idea of composing conditionally-normalized models on trees
is intuitive, the resulting models can be very complex, even for
simple models of RNA evolution, making (1) necessary.
Studies of related indel models have suggested that an implementation
of dynamic programming (DP) algorithms on the three-branch tree topology is sufficient to
draw samples from the posterior distribution of ancestral sequences on
more complex tree topologies, using Markov Chain Monte Carlo or MCMC
\cite{HolmesBruno2001,JensenHein2002,RedelingsSuchard2005},
suggesting that (2) and (3) are, in principle, sufficient for analyzing trees
relating many sequences.

% Benchmarking
We show that our algorithm produces a multiple-sequence grammar which is
much more compact than suggested by naive approaches to model
construction.  We provide analyses of the asymptotic complexities of
models constructed using our procedure and provide estimates of the
time and memory required to reconstruct the structures of several RNA
families for the case of a 3-taxon phylogeny, which we have
implemented in the program \indiegram. While by these estimates only
the smallest sequences currently fit into affordable memory, thereby
preventing us from applying our method to many problems of interest,
a simulation study suggests that we can hope to accurately reconstruct
ancestral structures over long evolutionary time, even in the presence
of structural divergence.

@@

Our problem statement is this:
%\vspace{.5\baselineskip}
%\fbox{
%\parbox{.8\textwidth}
\textbf{Given a phylogenetic tree relating several structured RNAs and a
  description of the evolution of a structured RNA
  along a single branch of the tree (in the form of a Pair SCFG),
  (1) find the corresponding
  phylogenetic multiple-sequence grammar and (2) use that grammar to
  reconstruct, {\em a posteriori}, the evolutionary histories of the
  RNAs.}
%}
%\vspace{.5\baselineskip}

We assume here that the phylogeny, including both the tree topology and branch lengths, is given.
This constraint somewhat arbitrarily confines the problem domain to the rather restricted case where we already have a phylogeny (and potentially a consensus structure),
and we seek only an alignment plus ancestral reconstruction.
It is important to note, however, that as far as the {\em model} is concerned, none of the constraints (phylogeny, structure or alignment) are final.
For example, the probabilistic framework readily allows us to compare likelihoods of two different phylogenetic trees, by constructing a composite transducer for each tree.
Thus, MCMC samplers for alignments could in principal be extended to phylogenies (albeit at a computational cost).
We revisit such MCMC samplers in the Discussion.

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
(We here use the term ``outgroup'' loosely, to denote the variable-length branch in a three-branch study,
where the other two branches are held at unit length.
Technically, the term ``outgroup'' is only correct if the outgroup's branch is longer than the other two branches.)

We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

@@

>>>>

We describe a ``transducer composition'' algorithm for extending
pairwise probabilistic models of RNA structural evolution to models of
multiple sequences related by a phylogenetic tree.  This algorithm
draws on formal models of computational linguistics as well as the
1985 protosequence algorithm of David Sankoff.  The output of the
composition algorithm is a multiple-sequence stochastic context-free
grammar.  We describe dynamic programming algorithms, which are robust
to null cycles and empty bifurcations, for parsing this grammar.
Example applications include structural alignment of non-coding RNAs,
propagation of structural information from an
experimentally-characterized sequence to its homologs, and inference
of the ancestral structure of a set of diverged RNAs.

@@

A number of leading methods for bioinformatics analysis of structural
RNAs use probabilistic grammars as models for pairs of homologous
RNAs.  We show that any such pairwise grammar can be extended to an
entire phylogeny by treating the pairwise grammar as a machine (a
``transducer'') that models a single ancestor-descendant relationship
in the tree, transforming one RNA structure into another.  In addition
to current applications such as RNA genefinding, homology detection,
alignment and secondary structure prediction, this should permit
probabilistic reconstruction of RNA genes ancestral to many
present-day sequences.  We describe algorithms, software
implementations, and an elementary benchmark of the dependence of the
accuracy of three-taxon RNA reconstruction on the outgroup branch
length.  In the Discussion we consider how the three-taxon RNA
alignment-reconstruction-folding algorithm, which is currently very
computationally-expensive, might be made more efficient so that larger
phylogenies could be considered.

@@

% (second part: multiple)
The second half of the phylogenetic modeling problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are
observed but those at internal nodes are not.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

@@

The main contributions of this paper are
(1) an algorithm that transforms a phylogenetic ensemble of pair
grammars, representing models on branches of a phylogenetic tree,
into a coherent, multiple-sequence SCFG,
(2) dynamic programming (DP) algorithms for performing inference under
this multiple-sequence SCFG, and
% (which appear to converge even if the SCFG contains null bifurcations and null cycles);
(3) freely-available software implementing algorithms (1) and (2) for
the simplified case of a three-taxon star-topology tree.
While the idea of composing conditionally-normalized models on trees
is intuitive, the resulting models can be very complex, even for
simple models of RNA evolution, making (1) necessary.
Studies of related indel models have suggested that an implementation
of dynamic programming (DP) algorithms on a three-taxon tree is sufficient to
draw samples from the posterior distribution of ancestral sequences on
more complex tree topologies, using Markov Chain Monte Carlo or MCMC
\cite{HolmesBruno2001,JensenHein2002,RedelingsSuchard2005},
suggesting that (2) and (3) are, in principle, sufficient for analyzing trees
relating many sequences.

% Benchmarking
We show that our algorithm produces a multiple-sequence grammar which is
much more compact than suggested by naive approaches to model
construction.  We provide analyses of the asymptotic complexities of
models constructed using our procedure and provide estimates of the
time and memory required to reconstruct the structures of several RNA
families for the case of a three-taxon phylogeny, which we have
implemented in the program \indiegram. While by these estimates only
the smallest sequences currently fit into affordable memory, thereby
preventing us from applying our method to many problems of interest,
a simulation study suggests that we can hope to accurately reconstruct
ancestral structures over long evolutionary time, even in the presence
of structural divergence.

@@

Our problem statement is this:
%\vspace{.5\baselineskip}
%\fbox{
%\parbox{.8\textwidth}
\textbf{Given a phylogenetic tree relating several structured RNAs and a
  description of the evolution of a structured RNA
  along a single branch of the tree (in the form of a Pair SCFG),
  (1) find the corresponding
  phylogenetic multiple-sequence grammar and (2) use that grammar to
  reconstruct, {\em a posteriori}, the evolutionary histories of the
  RNAs.}
%}
%\vspace{.5\baselineskip}
We assume here that the phylogeny, including both the tree topology
and branch lengths, is given.
% Ian, I moved the bits which you added here to the discussion; please
% edit per your discretion.

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
(We here use the term ``outgroup'' loosely to denote the
variable-length branch in our three-taxon study,
where the other two branches are held at unit length.)

We simulated the evolution of RNAs under the TKFST model along
three-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

@@

This paper focuses on the case where the tree topology is known, but
many of the methods which we describe below can be extended to the
more general case where none of the possible constraints (phylogeny,
structure or alignment) are final.  For example, the probabilistic
framework readily allows us to compare likelihoods of two different
phylogenetic trees by constructing a composite transducer for each
tree.  Thus, the MCMC samplers described above for alignments could,
in principal, be extended to phylogenies (albeit at a computational
cost).
